Imputation of Incomplete Multilevel Data with \proglang{R}

### 2023-10-27 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Add convergence plots

title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}. Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data. More complex models can be accomodated, but are outside the scope of this paper. 
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

@@

With incomplete clustered datasets we can distinguish between two types of missing data: sporadic missingness and systematic missingness [@resc13]. Sporadic missingness arises when variables are missing for some but not all of the units in a cluster [@buur18; @jola18]. For example, it is possible that test results are missing for several students in one or more classes. When all observations are missing within one or more clusters, data are said to be systematically missing. 
<!-- [TODO: Refer to Figure 1 and put interpretation in the figure caption. VMTdJ: "Mention in text that this is sporadic missingness?"] -->
<!-- TODO: Provide an example for one of the case studies below.] -->
<!-- TODO: add that indicator method doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low. -->

@@

Install non-CRAN packages if necessary:
```{r packages, eval=FALSE, include=TRUE}
devtools::install_github("amices/ggmice")
# devtools::install_github("hanneoberman/miceheckman")
```

@@

Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(2022)        # for reproducibility
library(mice)         # for imputation
library(miceadds)     # for additional imputation routines
library(ggmice)       # for incomplete/imputed data visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel parameter pooling
library(micemd)  # for imputation cf. heckman models
library(metamisc)     # for case study data
```
<!-- library(broom.mixed)  # for multilevel estimates -->

@@

@@

 <!-- The ICC in the incomplete data is `round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`. This tells us that the multilevel structure of the data should probably be taken into account. If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.-->

@@

<!-- [TODO: remove the broom.mixed output, use mitml only] -->

@@

```{r}
testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

@@

@@

Print the estimates:
```{r}
testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

@@

@@

The estimated effects in the complete data are visualized in Figure \ref{}.

```{r forest, echo=FALSE}
fits <-
  purrr::map_dfr(
    unique(impact$name),
    ~ {
      glm(
        mort ~ 1 + age + motor_score + pupil + ct,
        family = "binomial",
        data = filter(impact, name == .x)
      )
    } %>%
      broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
      .[, c("term", "estimate", "conf.low", "conf.high")] %>%
      cbind(name = as.character(.x), .)
  ) %>% filter(term %in% c("ctIII", "ctIV/V"))

n <- impact %>% group_by(name) %>% 
  tally()

fits %>% left_join(n, by = "name") %>% 
  ggplot(aes(y = name, x = estimate, color = term, linewidth = n)) +
  geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
  geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
  geom_point(shape = 20, position = position_dodge(width = 0.5)) +
  labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
  theme_classic() +
  scale_size(guide = 'none')
# TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
# TODO: missingness prop inside clusters
# TODO: make row height proportional to pattern frequency
```
<!-- We will use the following estimates as comparative truth in this tutorial: -->
```{r truth}
# fit <- glmer(mod, family = "binomial", data = impact) # fit the model
# tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
```

@@

```{r impact-fit, cache=TRUE}
fit <- imp %>% 
  with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
tidy(pool(fit))
as.mitml.result(fit)
# testEstimates(as.mitml.result(fit))
```

@@

Data are simulated and included in the `miceheckman` package. We will use the following variables:

>>>>

title:
  formatted: "Imputation of Incomplete Multilevel Data with \\proglang{R}"
  plain:     "Imputation of Incomplete Multilevel Data with R"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}. Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data. More complex models can be accomodated, but are outside the scope of this paper. 
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

## Multilevel data

@@

With incomplete clustered datasets we can distinguish between two types of missing data: sporadic missingness and systematic missingness [@resc13]. Sporadic missingness arises when variables are missing for some but not all of the units in a cluster [@buur18; @jola18]. For example, it is possible that test results are missing for several students in one or more classes. When all observations are missing within one or more clusters, data are said to be systematically missing. Sporadic missingness is visualized in Figure XYZ. 
<!-- [TODO: Refer to Figure 1 and put interpretation in the figure caption. VMTdJ: "Mention in text that this is sporadic missingness?"] -->
<!-- TODO: Provide an example for one of the case studies below.] -->
<!-- TODO: add that indicator method doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low. -->

@@

@@

Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(123)         # for reproducibility
library(mice)         # for imputation
library(miceadds)     # for additional imputation routines
library(ggmice)       # for incomplete/imputed data visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel parameter pooling
library(micemd)       # for case study data and imputation cf. heckman models
library(metamisc)     # for case study data
library(broom.mixed)  # for multilevel estimates
```

TODO: add table with predictor matrix values

- -2 = cluster variable 
- 1 = overall effect
- 3 = overall + group-level effect 
- 4 = individual-level (random) and group-level (fixed) effect

@@

The estimated effects in the complete data are presented in Table XYZ. We consider the associations in the full data set to be the true associations.

@@

The ICC in the incomplete data is `round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`. This tells us that the multilevel structure of the data should probably be taken into account. If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.

@@

Evaluate the convergence of the algorithm:
```{r pop-ignored-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
plot_trace(imp)
```
<!-- [TODO: remove the broom.mixed output, use mitml only] -->

@@

```{r pop-print, eval=FALSE}
testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

@@

Evaluate the convergence of the algorithm:
```{r pop-predictor-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
plot_trace(imp)
```

@@

Print the estimates:
```{r, eval=FALSE}
testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

@@

Evaluate the convergence of the algorithm:
```{r pop-multilevel-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
plot_trace(imp)
```

@@

The estimated effects in the complete data are presented in Table XYZ.

<!-- visualized in Figure \ref{}. -->

```{r forest, echo=FALSE}
# fits <-
#   purrr::map_dfr(
#     unique(impact$name),
#     ~ {
#       glm(
#         mort ~ 1 + age + motor_score + pupil + ct,
#         family = "binomial",
#         data = filter(impact, name == .x)
#       )
#     } %>%
#       broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
#       .[, c("term", "estimate", "conf.low", "conf.high")] %>%
#       cbind(name = as.character(.x), .)
#   ) %>% filter(term %in% c("ctIII", "ctIV/V"))
# 
# n <- impact %>% group_by(name) %>% 
#   tally()
# 
# fits %>% left_join(n, by = "name") %>% 
#   ggplot(aes(y = name, x = estimate, color = term, linewidth = n)) +
#   geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
#   geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
#   geom_point(shape = 20, position = position_dodge(width = 0.5)) +
#   labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
#   theme_classic() +
#   scale_size(guide = 'none')
# # TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
# # TODO: missingness prop inside clusters
# # TODO: make row height proportional to pattern frequency
```
<!-- We will use the following estimates as comparative truth in this tutorial: -->
```{r truth}
# fit <- glmer(mod, family = "binomial", data = impact) # fit the model
# tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
```

@@

Evaluate the convergence of the algorithm:
```{r impact-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
plot_trace(imp)
```

Analyze the imputed data:
```{r impact-fit, cache=TRUE}
fit <- imp %>% 
  with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
tidy(pool(fit))
as.mitml.result(fit)
# testEstimates(as.mitml.result(fit))
```

@@

Data are simulated and included in the `micemd` package. We will use the following variables:

### 2023-10-27 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Add convergence plots

\author{
Hanne I. Oberman\\Methodology and Statistics\\
Utrecht University \And Johanna Muñoz\\Julius Center for Health Sciences
and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
for Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \And Gerko Vink\\Methodology and Statistics\\
Utrecht University \AND Valentijn M. T. de Jong\\Julius Center for
Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands\\
Data Analytics and Methods Task Force,\\
European Medicines Agency,\\
Amsterdam, The Netherlands
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne I. Oberman, Johanna Muñoz, Thomas P. A. Debray, Gerko
Vink, Valentijn M. T. de Jong}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{Multilevel \pkg{mice}}

\Abstract{
This tutorial illustrates the imputation of incomplete multilevel data
with the \proglang{R} packackage \pkg{mice}.
}

@@

@@

With incomplete clustered datasets we can distinguish between two types
of missing data: sporadic missingness and systematic missingness
\citep{resc13}. Sporadic missingness arises when variables are missing
for some but not all of the units in a cluster \citep{buur18, jola18}.
For example, it is possible that test results are missing for several
students in one or more classes. When all observations are missing
within one or more clusters, data are said to be systematically missing.

@@

Install non-CRAN packages if necessary:

\begin{CodeChunk}
\begin{CodeInput}
R> devtools::install_github("amices/ggmice")
R> # devtools::install_github("hanneoberman/miceheckman")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)        # for reproducibility
R> library(mice)         # for imputation
R> library(miceadds)     # for additional imputation routines
R> library(ggmice)       # for incomplete/imputed data visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(mitml)        # for multilevel parameter pooling
R> library(micemd)  # for imputation cf. heckman models
R> library(metamisc)     # for case study data
\end{CodeInput}
\end{CodeChunk}

@@

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> testEstimates(as.mitml.result(fit), extra.pars = TRUE)
\end{CodeInput}
\begin{CodeOutput}

Call:

testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)

Final parameter estimates and inferences obtained from 5 imputed data sets.

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     5.012     0.295    16.994     4.362     0.000    22.587     0.969 
sex             0.695     0.251     2.768     4.287     0.047    28.390     0.975

                            Estimate 
Intercept~~Intercept|school    0.266 
Residual~~Residual             1.035 
ICC|school                     0.208

Unadjusted hypothesis test as appropriate in larger samples.
\end{CodeOutput}
\end{CodeChunk}

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> testEstimates(as.mitml.result(fit), extra.pars = TRUE)
\end{CodeInput}
\begin{CodeOutput}

Call:

testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)

Final parameter estimates and inferences obtained from 5 imputed data sets.

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     4.915     0.217    22.642     4.926     0.000     9.110     0.926 
sex             0.975     0.283     3.444     4.250     0.024    32.504     0.978

                            Estimate 
Intercept~~Intercept|school    0.351 
Residual~~Residual             1.153 
ICC|school                     0.233

Unadjusted hypothesis test as appropriate in larger samples.
\end{CodeOutput}
\end{CodeChunk}

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> testEstimates(as.mitml.result(fit), extra.pars = TRUE)
\end{CodeInput}
\begin{CodeOutput}

Call:

testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)

Final parameter estimates and inferences obtained from 5 imputed data sets.

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     5.011     0.410    12.222     4.226     0.000    35.955     0.980 
sex             0.928     0.381     2.434     4.168     0.069    48.221     0.985

                            Estimate 
Intercept~~Intercept|school    0.313 
Residual~~Residual             1.428 
ICC|school                     0.188

Unadjusted hypothesis test as appropriate in larger samples.
\end{CodeOutput}
\end{CodeChunk}

@@

The estimated effects in the complete data are visualized in Figure
\ref{}.

\begin{CodeChunk}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/forest-1} \end{center}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- imp %>% 
+   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
R> tidy(pool(fit))
\end{CodeInput}
\begin{CodeOutput}
                     term    estimate   std.error  statistic      p.value
1             (Intercept) -2.35199614 0.340174191  -6.914093 4.993068e-12
2                    type -0.41267936 0.180271952  -2.289204 2.208652e-02
3                     age  0.03049024 0.001570154  19.418637 1.233010e-81
4 as.factor(motor_score)2 -0.66765325 0.068737838  -9.713038 3.478858e-22
5 as.factor(motor_score)3 -1.05520488 0.070218678 -15.027410 2.536132e-50
6 as.factor(motor_score)4 -1.51238928 0.072304221 -20.917026 1.857373e-90
7                   pupil  0.48421506 0.038982924  12.421209 6.781796e-17
8                      ct  0.43474556 0.029968441  14.506779 2.341875e-36
             b          df dfcom         fmi      lambda m         riv
1 5.301620e-04 10113.60453 11013 0.005694385 0.005497777 5 0.005528170
2 4.905995e-05 10892.82806 11013 0.001994781 0.001811557 5 0.001814845
3 3.333628e-08  6323.59900 11013 0.016537098 0.016226102 5 0.016493731
4 4.194026e-05  8322.17049 11013 0.010889419 0.010651742 5 0.010766424
5 5.136864e-05  7631.13821 11013 0.012760548 0.012501842 5 0.012660117
6 1.403437e-04  2830.62134 11013 0.032897234 0.032214161 5 0.033286456
7 3.571667e-04    49.96887 11013 0.309144148 0.282035215 5 0.392825972
8 8.586575e-05   294.70102 11013 0.120676307 0.114728924 5 0.129597506
          ubar
1 1.150823e-01
2 3.243910e-02
3 2.425379e-06
4 4.674562e-03
5 4.869020e-03
6 5.059488e-03
7 1.091068e-03
8 7.950686e-04
\end{CodeOutput}
\begin{CodeInput}
R> as.mitml.result(fit)
\end{CodeInput}
\begin{CodeOutput}
[[1]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10495.423 10561.192 -5238.712 10477.423     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2843  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.37194                 -0.41015                  0.03052  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65802                 -1.04611                 -1.51245  
                  pupil                       ct  
                0.50405                  0.42496  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[2]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
     AIC      BIC   logLik deviance df.resid 
10500.88 10566.65 -5241.44 10482.88    11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2917  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.37718                 -0.41511                  0.03067  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66935                 -1.05211                 -1.49429  
                  pupil                       ct  
                0.49013                  0.43835  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[3]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10505.026 10570.795 -5243.513 10487.026     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2908  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.32327                 -0.42365                  0.03023  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.67143                 -1.05776                 -1.51039  
                  pupil                       ct  
                0.49756                  0.42474  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[4]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10519.511 10585.280 -5250.755 10501.511     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2961  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.33576                 -0.40872                  0.03039  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66477                 -1.05453                 -1.51860  
                  pupil                       ct  
                0.45928                  0.44419  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[5]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10522.038 10587.807 -5252.019 10504.038     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2955  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.35183                 -0.40576                  0.03064  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.67469                 -1.06551                 -1.52622  
                  pupil                       ct  
                0.47006                  0.44148  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

@@

Data are simulated and included in the \texttt{miceheckman} package. We
will use the following variables:

>>>>

\author{
Hanne I. Oberman\\Methodology and Statistics\\
Utrecht University \And Johanna Muñoz\\Julius Center for Health Sciences
and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
for Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \And Gerko Vink\\Methodology and Statistics\\
Utrecht University \AND Valentijn M. T. de Jong\\Julius Center for
Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands\\
Data Analytics and Methods Task Force,\\
European Medicines Agency,\\
Amsterdam, The Netherlands
}
\title{Imputation of Incomplete Multilevel Data with \proglang{R}}

\Plainauthor{Hanne I. Oberman, Johanna Muñoz, Thomas P. A. Debray, Gerko
Vink, Valentijn M. T. de Jong}
\Plaintitle{Imputation of Incomplete Multilevel Data with R}
\Shorttitle{Multilevel \pkg{mice}}

\Abstract{
This tutorial illustrates the imputation of incomplete multilevel data
with the \proglang{R} packackage \pkg{mice}. Our scope is only simple
multilevel models, to show how imputation can yield less biased
estimates from incomplete clustered data. More complex models can be
accomodated, but are outside the scope of this paper.
}

@@

\hypertarget{multilevel-data}{%
\subsection{Multilevel data}\label{multilevel-data}}

@@

With incomplete clustered datasets we can distinguish between two types
of missing data: sporadic missingness and systematic missingness
\citep{resc13}. Sporadic missingness arises when variables are missing
for some but not all of the units in a cluster \citep{buur18, jola18}.
For example, it is possible that test results are missing for several
students in one or more classes. When all observations are missing
within one or more clusters, data are said to be systematically missing.
Sporadic missingness is visualized in Figure XYZ.

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(123)         # for reproducibility
R> library(mice)         # for imputation
R> library(miceadds)     # for additional imputation routines
R> library(ggmice)       # for incomplete/imputed data visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(mitml)        # for multilevel parameter pooling
R> library(micemd)       # for case study data and imputation cf. heckman models
R> library(metamisc)     # for case study data
R> library(broom.mixed)  # for multilevel estimates
\end{CodeInput}
\end{CodeChunk}

TODO: add table with predictor matrix values

\begin{itemize}
\tightlist
\item
  -2 = cluster variable
\item
  1 = overall effect
\item
  3 = overall + group-level effect
\item
  4 = individual-level (random) and group-level (fixed) effect
\end{itemize}

@@

The ICC in the incomplete data is
\texttt{round(icc(popular\ \textasciitilde{}\ as.factor(school),\ data\ =\ na.omit(popmis)),\ 2)}.
This tells us that the multilevel structure of the data should probably
be taken into account. If we don't, we'll may end up with incorrect
imputations, biasing the effect of the clusters towards zero.

@@

Evaluate the convergence of the algorithm:

\begin{CodeChunk}
\begin{CodeInput}
R> plot_trace(imp)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-ignored-conv-1} \end{center}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> testEstimates(as.mitml.result(fit), extra.pars = TRUE)
\end{CodeInput}
\end{CodeChunk}

@@

Evaluate the convergence of the algorithm:

\begin{CodeChunk}
\begin{CodeInput}
R> plot_trace(imp)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-predictor-conv-1} \end{center}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> testEstimates(as.mitml.result(fit), extra.pars = TRUE)
\end{CodeInput}
\end{CodeChunk}

@@

Evaluate the convergence of the algorithm:

\begin{CodeChunk}
\begin{CodeInput}
R> plot_trace(imp)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-multilevel-conv-1} \end{center}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> testEstimates(as.mitml.result(fit), extra.pars = TRUE)
\end{CodeInput}
\end{CodeChunk}

@@

The estimated effects in the complete data are presented in Table XYZ.

@@

Evaluate the convergence of the algorithm:

\begin{CodeChunk}
\begin{CodeInput}
R> plot_trace(imp)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-conv-1} \end{center}

Analyze the imputed data:

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- imp %>% 
+   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
R> tidy(pool(fit))
\end{CodeInput}
\begin{CodeOutput}
                     term    estimate   std.error  statistic      p.value
1             (Intercept) -2.29521346 0.349767178  -6.562118 7.886872e-11
2                    type -0.41750657 0.180984169  -2.306868 2.108127e-02
3                     age  0.03038268 0.001579024  19.241433 2.785453e-78
4 as.factor(motor_score)2 -0.65834010 0.068555350  -9.603045 9.808985e-22
5 as.factor(motor_score)3 -1.05935395 0.070432441 -15.040711 4.685873e-50
6 as.factor(motor_score)4 -1.51332652 0.072497350 -20.874232 1.222074e-88
7                   pupil  0.45660452 0.043112821  10.590922 2.255987e-10
8                      ct  0.43389476 0.034110953  12.720101 1.856548e-15
             b          df dfcom         fmi      lambda m         riv
1 5.551031e-03  1194.39341 11013 0.056029215 0.054449867 5 0.057585384
2 9.060570e-05 10652.43106 11013 0.003506444 0.003319369 5 0.003330424
3 6.016742e-08  3298.53424 11013 0.029546025 0.028957788 5 0.029821348
4 3.222589e-05  9216.80652 11013 0.008443314 0.008228174 5 0.008296439
5 8.436583e-05  5080.45332 11013 0.020793493 0.020408089 5 0.020833256
6 1.633007e-04  2263.14149 11013 0.038133867 0.037284215 5 0.038728164
7 6.411714e-04    23.25984 11013 0.458579946 0.413944862 5 0.706324089
8 3.096973e-04    39.00604 11013 0.351802055 0.319397049 5 0.469285431
          ubar
1 1.156758e-01
2 3.264654e-02
3 2.421115e-06
4 4.661165e-03
5 4.859490e-03
6 5.059905e-03
7 1.089310e-03
8 7.919204e-04
\end{CodeOutput}
\begin{CodeInput}
R> as.mitml.result(fit)
\end{CodeInput}
\begin{CodeOutput}
[[1]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10498.989 10564.757 -5240.494 10480.989     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2895  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.39617                 -0.40607                  0.03021  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65393                 -1.05692                 -1.49555  
                  pupil                       ct  
                0.48407                  0.45095  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[2]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10502.536 10568.305 -5242.268 10484.536     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2926  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
                -2.3350                  -0.4268                   0.0301  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
                -0.6647                  -1.0565                  -1.5065  
                  pupil                       ct  
                 0.4827                   0.4481  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[3]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10574.673 10640.442 -5278.337 10556.673     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2921  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.23276                 -0.41737                  0.03068  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66414                 -1.07209                 -1.52198  
                  pupil                       ct  
                0.44804                  0.40791  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[4]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10548.019 10613.788 -5265.009 10530.019     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2953  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.29768                 -0.41020                  0.03058  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65615                 -1.04746                 -1.52793  
                  pupil                       ct  
                0.43858                  0.43668  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[5]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10562.267 10628.035 -5272.133 10544.267     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2949  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.21445                 -0.42713                  0.03034  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65280                 -1.06379                 -1.51469  
                  pupil                       ct  
                0.42960                  0.42587  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

@@

Data are simulated and included in the \texttt{micemd} package. We will
use the following variables:
